Tidy Data

### 2014-01-06 reviews/response.md

# Reviewer 1

> Details of writing: There is opportunity for improvement and for 
> strengthening the exposition at a local level – in sentences and phrases. 
> This reviewer has just read Wickham's 2011 paper in the same journal on
> "Split-Apply-Combine", and the writing in that paper is good and can serve 
> as a model for polishing the present paper. The differences between the two 
> are in the details, and a thorough editing and tightening of the present 
> manuscript would almost surely pay dividends. Without trying to be 
> comprehensive, I will illustrate the kinds of small changes (including 
> just fixing typos) which taken together would have a positive effect. (R1)

> Abstract: "take in and take out" "input and output"
> p. 1 up 6: "subset" -> "component"

> p. 1 last full sentence: "The reorganization makes...because it conforms to 
> a standard that facilitates well an initial exploration and analysis of the data; you don't..."

> p. 2 line 2: Is "reformulating" an approximation for "munging"?
￼
> p. 2 line 9: -> "an extension..."

> p. 2, 5 up from subtitle: "...techniques with real examples." -- could do 
> nicely.

> p. 2, 1 up from subtitle "...misses and what other approaches might be 
> fruitful to pursue."

> p. 3 2/3 down "...were were..."

> p. 3 bottom. Consider having a new bold subtitle as a guidepost for the 
> reader: Defining tidy data:

> p. 10, up 2. Sentence is garbled.

> p. 11 bottom: (left) and (right) may not be the apt descriptors...

> p. 12, line 9: comma after file name?

> p. 12: This reviewer would prefer to avoid "hopefully" about 13 lines down. 
> More substantively, the sentence is garbled.

> p. 13 middle "...by the by preposition." Should second "by' be bolded? Or in 
> quotes" Same question arises again.

> p. 14 last paragraph. A difficulty is identified; does the author suggest 
> a solution? Reader isn't clear where we are left as we enter the Case Study.

> p. 17 middle -> " diseases we work with have..."

> p. 18 nice plots in an interesting context!

> p. 19 last line "...seem like they should ..." -> "may"

> p. 20 very effective use of exploratory data analysis

> p. 21 down 13: might be clearer if "...an efficient equivalent to join." 
> – bolding or perhaps quotes.

> This reviewer notes (in a spirit of collegiality) that the best writers about 
> data analysis (from a previous generation) – Fred Mosteller, John Tukey,
> David Hoaglin, Paul Velleman, Frank Anscombe (as illustrations) – correctly 
> used data as plural form (and occasionally datum as singular). From that 
> perspective, "the data is ..." grates more than a little ... but 
> regrettably this is a battle that is now lost. Sigh... :) :)

# Reviewer 2

> While the topic is important and the message is clear, I found the tone of
> the paper to be too conversational. The author overuses contractions,
> ambiguous pronouns ("It's often said..."), and casually addresses the reader 
> in second-person form. (R2)

> The author also switches between the use of italics, boldface, quotes, and 
> the LaTeX verbatim environment constantly, which tends to be distracting 
> at times. (R2)

> - The author uses "colvars" in italics when first naming the terminology, and 
> then subsequently mentions the Pew dataset contains one colvar, without 
> italics. (R2)

> - The author does not use formatting on the column names in the caption for 
> Table 9 but does use \verbatim to reference column names in Table 10's 
> caption. (R2)

> On page 13, middle of the page, it would be helpful to use quotes around 
> "by" where applicable, as "modified by the by preposition" does not read well.
> (R2)

> The author overuses colons. In some cases, I found hyphens to be more 
> appropriate, and in others, I would think semicolons might be more suitable.
> Some tables also have unreadable characters in them, such as the degree sign 
> in Table 12's top panel (row 6's artist). Same issue with Table 3's 5th 
> religion. Table 12's caption also describes there being a left and right 
> dataset, whereas the subtables are clearly placed above and below.
> (R2)

> The last sentence in the first paragraph of the discussion needs to be 
> rephrased for clarity.
> (R2)

> - Abstract: "variables are stored in columns, observations in rows, and a 
> single type of..." -- this list isn't logically homogeneous.

> - p.5, 3.1 -- "The Pew Center is an American ... that collects data on 
> attitudes to topics RANGING from religion to the internet"

> - "Billboard" should always be capitalized as it is a proper noun.

> - p.13, section 4.2 - "Tidy visualization tools ONLY NEED to be..."

> - p.16, 4-th line from bottom: "Next, we work out THE overall..."

> - p.16, 2nd line from the bottom: "Then finally, WE join.."

> - p. 17, "To ensure that the diseaseS we work with..."

> - p.19, "The causes of death fall INTO three main groups: ... " There should 
> be a hyphen between "transportation" and "related".

> - p. 19, 2nd paragraph of discussion: "This makes it easy"; remove "is".

> - p. 21, last sentence of first paragraph: "and A BETTER KNOWLEDGE OF how 
> we can best design tools..."

> - p. 21, last paragraph before Acknowledgements "verifying experimental 
> design, AND filling in..."

> - Author information, last page "Adjunct ASSISTANT Professor"

# Reviewer 3

> The author is obviously intimately familiar with the tools he's using in 
> this paper, and he should remember that the readers of this paper are not 
> nearly as familiar with them as he is. (R3)

> Section 2: "When using tables for communication, this so-called 
> "Alabama first" (Wainer 2000) ordering should be abandoned, and replaced 
> with an ordering based on a meaningful variable."
> 
> There doesn't seem to be any justification or explanation for this statement.
> (R3)

> Section 3.1: Since most readers of this paper will want to apply these ideas 
> using R, I think it would be a good idea to show how an R datasets containing 
> the Pew and Billboard data would look, as well as the R code that would be 
> used to melt the datasets. This is especially true for the Billboard data, 
> because additional cleaning was performed. Similar comments apply to the 
> other datasets in the paper. (R3)

> Section 3.5: I feel the examples in this section need to be explained in 
> more detail. The referenced external links contain a large number of files, 
> and it's not really clear which files are relevant to the discussion in this 
> section. (R3)

> Section 4.1: "Compare this to the difficulty of combining datasets stored in 
> arrays; these typically require painstaking alignment before matrix 
> operations can be used, and errors can be very hard to detect."
> Would anyone seriously suggest using matrices to store data which would be 
> combined with other data sets? The use of an index to combine multiple 
> datasets doesn't really seem to be related to tidy data. (R3)

> Section 4.3: I think readers of this paper might find examples from stata 
> more relevant than those from spss. (R3)

> Section 5: I was not able to download the data for the case study from 
> https://raw.github.com/hadley/tidy-data/master/case-study/deaths.rdata; 
> I recieved a 403 Forbidden error. (R3)

> I think there should be a more thorough discussion explaining the goal of 
> analyzing this data set. I'm not familiar with the term "time course", and 
> the single sentence "The case study uses individual-level mortality data f
> rom Mexico, with the goal of finding causes of death that have notably 
> different time patterns within a day." really didn't clarify the goal 
> sufficiently for me. Perhaps an example of an unusual time course would help 
> to clarify things. (R3)

### 2014-01-07 reviews/response.md

I really appreciate the thoughtful comments of all the reviewers. I've worked
hard to make the paper easier to understand and the reviewer's comments were
extremely useful.  A point-by-point summary of the comments and my rebuttal
is below. Minor comments fixed without needing further discussion are listed
at the end.

@@

# Reviewer 1

> Details of writing: There is opportunity for improvement and for 
> strengthening the exposition at a local level – in sentences and phrases. 
> This reviewer has just read Wickham's 2011 paper in the same journal on
> "Split-Apply-Combine", and the writing in that paper is good and can serve 
> as a model for polishing the present paper. The differences between the two 
> are in the details, and a thorough editing and tightening of the present 
> manuscript would almost surely pay dividends. Without trying to be 
> comprehensive, I will illustrate the kinds of small changes (including 
> just fixing typos) which taken together would have a positive effect. (R1)

> p. 1 last full sentence: "The reorganization makes...because it conforms to 
> a standard that facilitates well an initial exploration and analysis of the 
> data; you don't..."

I'm not quite sure what the reviewer meant here, but I've rewritten the 
paragraph for clarity.

> p. 2 line 2: Is "reformulating" an approximation for "munging"?

Yes, but I think mungling better conveys the inelegance of this operation :)

@@

> p. 3 bottom. Consider having a new bold subtitle as a guidepost for the 
> reader: Defining tidy data:

I've completely reorganised this section and it should be much more clear.

@@

> p. 14 last paragraph. A difficulty is identified; does the author suggest 
> a solution? Reader isn't clear where we are left as we enter the Case Study.

> p. 21 down 13: might be clearer if "...an efficient equivalent to join." 
> – bolding or perhaps quotes.

I've removed those comments because I know no longer think that they a true.

@@

> Section 4.3: I think readers of this paper might find examples from stata 
> more relevant than those from spss. (R3)

> Section 5: I was not able to download the data for the case study from 
> https://raw.github.com/hadley/tidy-data/master/case-study/deaths.rdata; 
> I recieved a 403 Forbidden error. (R3)

> I think there should be a more thorough discussion explaining the goal of 
> analyzing this data set. I'm not familiar with the term "time course", and 
> the single sentence "The case study uses individual-level mortality data f
> rom Mexico, with the goal of finding causes of death that have notably 
> different time patterns within a day." really didn't clarify the goal 
> sufficiently for me. Perhaps an example of an unusual time course would help 
> to clarify things. (R3)

# Minor problems

> Abstract: "take in and take out" "input and output"

> p. 1 up 6: "subset" -> "component"

> p. 2 line 9: -> "an extension..."

### 2014-01-07 tidy-data.tex
Changes from reviewer 1

\documentclass[article]{jss}

@@

%% an abstract and keywords
\Abstract{
A huge amount of effort is spent cleaning data to get it ready for data analysis, but there has been little research on how to make data cleaning as easy and effective as possible. This paper tackles a small, but important, subset of data cleaning: data ``tidying''. Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: variables are stored in columns, observations in rows, and a single type of experimental unit per dataset. This framework makes it easy to tidy messy datasets because only a small set of tools are needed to deal with a wide range of un-tidy datasets. This structure also makes it easier to develop tidy tools for data analysis, tools that both take in and put out tidy datasets. The advantages of a consistent data structure and matching tools are demonstrated with a case study free from mundane data manipulation chores.
}
\Keywords{data cleaning, data tidying, relational databases, \proglang{R}}
\Plainkeywords{data cleaning, data tidying, relational databases, R}

@@

The principles of tidy data provide a standard way of organising the data values within a dataset. This makes initial data cleaning easier because you have a \emph{standard} that you can start with: you don't need to reinvent the wheel for each new dataset. It also makes it easier to develop data analysis tools that work together seamlessly. Current tools are often inconsistent, so you need to spend time munging the output from one tool into the input for another. Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not the uninteresting details of data structure.

The principles of tidy data are closely tied to relational databases and Codd's relational algebra \citep{codd:1990}, but are framed in a language more familiar to statisticians. Computer scientists have also contributed much to the study of data cleaning. For example, \citet{lakshmanan:1996} defines a extension to SQL allowing it to operate on messy datasets, \citet{raman:2001} provides a framework for cleaning datasets, and \citet{kandel:2011} develops an interactive tool that makes data cleaning easy, automatically creating the code to clean datasets with a friendly user interface. These tools are useful, but are presented in language foreign to most statisticians, fail to give much advice on how datasets should be structured, and lack connections to the tools of data analysis.

@@

The paper proceeds as follows. Section~\ref{sec:defining} begins by defining the three characteristics that make a dataset tidy. Most real world datasets are not tidy, so Section~\ref{sec:tidying} describes the operations needed to make messy datasets tidy, and illustrates the techniques with a range of real-life examples. Section~\ref{sec:tidy-tools} defines tidy tools, tools that input and output tidy datasets, and discusses how together tidy data and tidy tools make it easier to do a data analysis. These principles are illustrated with a small case study in Section~\ref{sec:case-study}. Section~\ref{sec:discussion} concludes with a discussion of what this framework misses and other approaches that might be fruitful to pursue.

@@

Like families, tidy datasets are all alike, but every messy dataset is messy in its own way. Here we will define the essence of tidy data, and provide important vocabulary for describing datasets.

Statistical datasets usually come in the form of a rectangular table, made up of rows and columns. Tables are usually labelled with column names, and sometimes with row names. The table is a collection of data values, typically either numeric (if quantitative) or character (if qualitative). Values are grouped into variables, measurements of a single quantifiable property. Multiple measurements made on the same observational unit form an observation.

There is a little circularity in the definition, which is resolved by defining an observational unit as the smallest constituent atom of the experimental unit. For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at each time point. In an agricultural experiment, the experimental units might be the cages given different diets, but the observational units are the pigs whose individual weights are measured. This definition of observational unit leads to variables which are physical properties, like blood pressure and weight, not conditional properties like blood pressure at time 1, or weight of pig 2. Structure in the variable names (`trtA', `trtB', `trtC'; `jan1', `feb1', `mar1') is difficult to exploit, and suggests that the names are values of other variables.

% Experimental units are usually independent, and the statistical challenge in an analysis is capturing the dependence of observational units within an experimental unit. Failing to do so correctly leads to problems like pseudoreplication \citep{hurlbert:1984}.

Table~\ref{tbl:preg-raw} illustrates a data format commonly seen in the wild. Here it is used to store information about an imaginary experiment with two treatments. There are two columns and three rows (plus column and row names), defining three variables (name, treatment, result) and either five or six observations, depending on how you think about the missing value.

\begin{table}[htbp]
  \centering
  \begin{tabular}{lrr}
  \toprule
               & treatmentA & treatmentB \\
  \midrule
  John Smith   &      ---  &            5 \\
  Jane Doe     &        1  &            4 \\
  Mary Johnson &        2  &            3 \\
  \bottomrule
  \end{tabular}
  \caption{Typical presentation dataset.}
  \label{tbl:preg-raw}
\end{table}

It's not always obvious what qualifies as a variable, and a variable in one analysis may be a value in another. For example, if the columns in the example were were height and weight we would have been happy to call them variables. If the columns were height and width, it would be less clear cut, as we might think of height and weight as values of a dimension variable. If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud. A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g., average of group a vs.\ average of group b) than between columns. If you find yourself fighting these constraints, you may not have correctly identified the appropriate variables for your problem.

In a given analysis, there may also be multiple types of observational unit. For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).

A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:

@@

Table~\ref{tbl:preg-tidy} shows the tidy version of Table~\ref{tbl:preg-raw}. Now each row represents an observation, the result of one treatment on one person, and each column is a variable. We could also choose to drop the row representing the result of John Smith's treatment a, but we don't, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.

\begin{table}[htbp]
  \centering
  \begin{tabular}{lrr}
    \toprule
    name &    treatment    & result  \\
    \midrule
    Jane Doe      & a & 1 \\
    Jane Doe      & b & 4 \\
    John Smith    & a & --- \\
    John Smith    & b & 5 \\
    Mary Johnson  & a & 2 \\
    Mary Johnson  & b & 3 \\
    \bottomrule
  \end{tabular}
  \caption{Tidied data.}
  \label{tbl:preg-tidy}
\end{table}

@@

  \caption{(a) Molten weather dataset.  This is almost tidy, but the \code{element} column contains names of variables, not values. Missing values are dropped to conserve space. (b) Tidy weather dataset. Each row representation the meteorological measurements for a single day.  There are two measured variables, minimum (\code{tmin}) and maximum (\code{tmax}) temperature; all other variables are fixed.}
  \label{tbl:weather-clean}
\end{table}

@@

\begin{table}
  \centering
  \input{data/billboard-song.tex}\hspace{2em}%
  \input{data/billboard-rank.tex}

@@

It's also common to find data values about a single type of observational unit spread out over multiple tables, or multiple files. These tables are often split up by another variable, so each file represents a single year, person, or location. As long as the format for individual records is consistent, this is an easy problem to fix: read in all files, add a new column that records the original file name (because the filename is often the value of an important variable), then combine all tables into a single table. Once you have a single table, you can perform additional tidying as needed. An example of this type of cleaning can be found at \url{https://github.com/hadley/data-baby-names} which takes 129 yearly baby name tables provided by the US Social Security Administration and combines them into a single file.

A more complicated situation occurs when the dataset structure has changed over time: different variables, variable names, data storage formats, or different conventions for missing values. This may require each file to tidied individually (or hopefully in small groups) and then combined together. An example of this type of tidying is illustrated in \url{https://github.com/hadley/data-fuel-economy}, which shows the tidying of {\sc epa} fuel economy data for over 50,000 cars from 1978 to 2008. The raw data is available online, but each year is stored in a separate file and there are four major formats with many minor variations, making tidying this dataset a considerable challenge.

@@

The four verbs can be, and often are, modified by the by preposition. We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on. Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time. Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy. Base \proglang{R} possesses a \code{by} function, which is input-tidy, but not output-tidy, because it produces a list. The \code{ddply} function from the \pkg{plyr} package is a tidy alternative.

@@

Next we compute a distance between the time course of each cause of death and the overall time course. Here we use a simple mean squared deviation.  We also record the sample size, the total number of deaths from that cause. To ensure that the disease we work with have a decent amount of underlying data we'll only 
work with diseases with more than 50 deaths in total ($\sim$2/hour).

@@

I have found few principles to guide the design of tidy data, which must acknowledge both statistical and cognitive factors. To date, my work has been driven by my experience doing data analysis, connections to relational database design, and introspection on the tools of data analysis. The human factors, user-centred design, and human-computer interaction communities seem like they should be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic. In the future, I hope to apply methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to my work to build a better understanding of the cognitive side of data analysis, and how we can best design tools to support the process.

@@

Other formulations of tidy data are possible. For example, it would be possible to construct a set of tools for dealing with values stored in multidimensional arrays. This is a common data storage format for large biomedical datasets such as microarray or fMRI data. It is also necessary for many multivariate methods whose underlying theory is based on matrix manipulation. This array-tidy format would be compact and efficient, because there are many efficient tools for working with high-dimensional arrays, even when sparse, and the format would connect more closely with the mathematical basis of statistics. This approach is taken by the Pandas python data analysis library \citep{mckinney:2010}. The biggest challenge is to connect different datasets, requiring the development of an efficient equivalent to join, and to connect to existing modelling tools without an expensive re-expression in another format. Even more interestingly, we could consider tidy tools that can ignore the underlying data representation, automatically choosing between array-tidy and dataframe-tidy formats to optimise memory usage and performance.

>>>>

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

@@

%% an abstract and keywords
\Abstract{
A huge amount of effort is spent cleaning data to get it ready for data analysis, but there has been little research on how to make data cleaning as easy and effective as possible. This paper tackles a small, but important, component of data cleaning: data ``tidying''. Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: variables are stored in columns, observations in rows, and a single type of experimental unit per dataset. This framework makes it easy to tidy messy datasets because only a small set of tools are needed to deal with a wide range of un-tidy datasets. This structure also makes it easier to develop tidy tools for data analysis, tools that both input and output tidy datasets. The advantages of a consistent data structure and matching tools are demonstrated with a case study free from mundane data manipulation chores.
}
\Keywords{data cleaning, data tidying, relational databases, \proglang{R}}
\Plainkeywords{data cleaning, data tidying, relational databases, R}

@@

The principles of tidy data provide a standard way of organising the data values within a dataset. A standard makes initial data cleaning easier because you don't need to start from scratch each time and you can avoid reinventing the wheel. The tidy data standard has been designed to facilitate initial exploration and analysis of the data, and it makes it easier to develop data analysis tools that work together seamlessly. Current tools are often inconsistent, so you need to spend time munging the output from one tool into the input for another. Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not the uninteresting logistics of your data.

The principles of tidy data are closely tied to relational databases and Codd's relational algebra \citep{codd:1990}, but are framed in a language more familiar to statisticians. Computer scientists have also contributed much to the study of data cleaning. For example, \citet{lakshmanan:1996} defines an extension to SQL allowing it to operate on messy datasets, \citet{raman:2001} provides a framework for cleaning datasets, and \citet{kandel:2011} develops an interactive tool that makes data cleaning easy, automatically creating the code to clean datasets with a friendly user interface. These tools are useful, but they are presented in a language foreign to most statisticians, they fail to give much advice on how datasets should be structured, and they lack connections to the tools of data analysis.

@@

The paper proceeds as follows. Section~\ref{sec:defining} begins by defining the three characteristics that make a dataset tidy. Most real world datasets are not tidy, so Section~\ref{sec:tidying} describes the operations needed to make messy datasets tidy, and illustrates the techniques with a range of real examples. Section~\ref{sec:tidy-tools} defines tidy tools, tools that input and output tidy datasets, and discusses how together tidy data and tidy tools make it easier to do a data analysis. These principles are illustrated with a small case study in Section~\ref{sec:case-study}. Section~\ref{sec:discussion} concludes with a discussion of what this framework misses and what other approaches might be fruitful to pursue.

@@

Like families, tidy datasets are all alike, but every messy dataset is messy in its own way. Tidy datasets provide a standard way to connect the structure of a dataset to its semantics. I'll provide some standard vocabulary for describing the structure of a dataset, it's semantics, and then use those definitions to define tidy data.

\subsection{Data structure}

Most statistical datasets are rectangular tables made up of rows and columns. The columns almost always labelled, and rows are often labelled. Table~\ref{tbl:preg-raw} provides some data about an imaginary experiment in a format commonly seen in the wild. The table has two columns and three rows, and both rows and columns are labelled.

\begin{table}[htbp]
  \centering
  \input{data/preg-raw-1.tex}
  \caption{Typical presentation dataset.}
  \label{tbl:preg-raw-1}
\end{table}

There are many ways to structure the same underlying data. Figure~\ref{tbl:preg-raw-2} shows exactly the same data, but the rows and columns have been transposed. Describing the structure of a dataset is not enough, we also need some way to describe it's underlying semantics.

\begin{table}[htbp]
  \centering
  \input{data/preg-raw-2.tex}
  \caption{The same data as in Table~\ref{tbl:preg-raw-1} but structured differently.}
  \label{tbl:preg-raw-2}
\end{table}

\subsection{Data semantics}

A table is a collection of data \textbf{values}, typically either numbers (if quantitative) or strings (if qualitative). Values are grouped into \textbf{variables}, measurements of a single quantifiable property. Multiple measurements made on the same observational unit form an \textbf{observation}. An observational unit as the smallest constituent atom of the experimental unit. For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at each time point. In an agricultural experiment, the experimental units might be the cages given different diets, but the observational units are the pigs whose individual weights are measured. This definition of observational unit leads to variables which are physical properties, like blood pressure and weight, not conditional properties like blood pressure at time 1, or weight of pig 2.

The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (name, treatment, result), six observations (two for each person), and either five or six values, depending on how you think about the missing value.  We can make this structure more obvious by putting the variables in the columns and the observations in the rows, as in Figure~\ref{tbl:preg-tidy}.

\begin{table}[htbp]
  \centering
  \input{data/preg-tidy.tex}
  \caption{Tidied data.}
  \label{tbl:preg-tidy}
\end{table}

For a given dataset, it's usually easy to figure out what are observations and what are variables, but precisely defining variables and observations in general is surprisingly difficult. For example, if the columns in the Table~\ref{fig:preg-raw-1} were height and weight we would have been happy to call them variables. If the columns were height and width, it would be less clear cut, as we might think of height and weight as values of a dimension variable. If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud. A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g., average of group a vs.\ average of group b) than between columns.

Note that in a given analysis, there may be multiple types of observational unit. For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).

\subsection{Tidy data}

Tidy data is a standard way of mapping the semantics of a dataset to its structure. A dataset is \emph{messy} or \emph{tidy} depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:

@@

Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw} because each row represents an observation, the result of one treatment on one person, and each column is a variable. We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.

@@

  \caption{(a) Molten weather dataset.  This is almost tidy, but the \code{element} column contains names of variables, not values. Missing values are dropped to conserve space. (b) Tidy weather dataset. Each row represents the meteorological measurements for a single day.  There are two measured variables, minimum (\code{tmin}) and maximum (\code{tmax}) temperature; all other variables are fixed.}
  \label{tbl:weather-clean}
\end{table}

@@

\begin{table}
  \centering
  \input{data/billboard-song.tex}\hspace{1em}%
  \input{data/billboard-rank.tex}

@@

It's also common to find data values about a single type of observational unit spread out over multiple tables, or multiple files. These tables are often split up by another variable, so each file represents a single year, person, or location. As long as the format for individual records is consistent, this is an easy problem to fix:

\begin{enumerate}
  \item read the files into a list of tables,

\item for each table, add a new column that records the original file name (because the file name is often the value of an important variable), then

  \item combine all tables into a single table. 
\end{enumerate}

Once you have a single table, you can perform additional tidying as needed. An example of this type of cleaning can be found at \url{https://github.com/hadley/data-baby-names} which takes 129 yearly baby name tables provided by the US Social Security Administration and combines them into a single file.

A more complicated situation occurs when the dataset structure has changed over time. For example, the datasets contain different variables, the same variables have different names, different file formats are used, or there are different conventions for missing values. This may require each file to tidied individually (or, if you're lucky, in small groups) and then combined once tidied. An example of this type of tidying is illustrated in \url{https://github.com/hadley/data-fuel-economy}, which shows the tidying of {\sc epa} fuel economy data for over 50,000 cars from 1978 to 2008. The raw data is available online, but each year is stored in a separate file and there are four major formats with many minor variations, making tidying this dataset a considerable challenge.

@@

The four verbs can be, and often are, modified by the ``by'' preposition. We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on. Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time. Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy. Base \proglang{R} possesses a \code{by} function, which is input-tidy, but not output-tidy, because it produces a list. The \code{ddply} function from the \pkg{plyr} package is a tidy alternative.

@@

Next we compute a distance between the time course of each cause of death and the overall time course. Here we use a simple mean squared deviation.  We also record the sample size, the total number of deaths from that cause. To ensure that the diseases we work with have a decent amount of underlying data we'll only 
work with diseases with more than 50 deaths in total ($\sim$2/hour).

@@

I have found few principles to guide the design of tidy data, which must acknowledge both statistical and cognitive factors. To date, my work has been driven by my experience doing data analysis, connections to relational database design, and introspection on the tools of data analysis. The human factors, user-centred design, and human-computer interaction communities may be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic. In the future, I hope to apply methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to my work to build a better understanding of the cognitive side of data analysis, and how we can best design tools to support the process.

@@

Other formulations of tidy data are possible. For example, it would be possible to construct a set of tools for dealing with values stored in multidimensional arrays. This is a common data storage format for large biomedical datasets such as microarray or fMRI data. It is also necessary for many multivariate methods whose underlying theory is based on matrix manipulation. This array-tidy format would be compact and efficient, because there are many efficient tools for working with high-dimensional arrays, even when sparse, and the format would connect more closely with the mathematical basis of statistics. This approach is taken by the Pandas python data analysis library \citep{mckinney:2010}. Even more interestingly, we could consider tidy tools that can ignore the underlying data representation, automatically choosing between array-tidy and dataframe-tidy formats to optimise memory usage and performance.
